Comparison of several derivative-free global optimization algorithms\thanks{The study was supported by the Russian Science Foundation, project No 16-11-10150.}

### 2019-01-20 2019/MOTOR_Sovrasov/motor_paper.tex
MOTOR: add a draft of paper

%%%%%%%%%%%%%%%%%%%% author.tex
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
% sample root file for your "contribution" to a proceedings volume
%
% Use this file as a template for your own input.
%
%%%%%%%%%%%%%%%% Springer
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{svproc}
%
% RECOMMENDED
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%
%
\usepackage{graphicx}
\usepackage{marvosym}
\usepackage{amssymb}
\usepackage{cite}
\usepackage{subfig}
% to typeset URLs, URIs, and DOIs
\usepackage{url}
\usepackage{hyperref}
\def\UrlFont{\rmfamily}

\def\orcidID#1{\unskip$^{[#1]}$}
\def\letter{$^{\textrm{(\Letter)}}$}

\begin{document}
\mainmatter              % start of a contribution
%
\title{Comparison of several sequential and parallel derivative-free global optimization algorithms}
%
\titlerunning{Comparison of several optimization methods}  % abbreviated title (for running head)
%                                     also used for the TOC unless
%                                     \toctitle is used
%
\author{Vladislav Sovrasov\letter \and Semen Bevzuk}
%
\authorrunning{Vladislav Sovrasov et al.} % abbreviated author list (for running head)
%
%%%% list of authors for the TOC (use if author list has to be modified)
\tocauthor{Vladislav Sovrasov and Semen Bevzuk}
%
\institute{Lobachevsky State University of Nizhni Novgorod, Russia \\
  \email{sovrasov.vlad@gmail.com, semen.bevzuk@gmail.com}
}

\maketitle              % typeset the title of the contribution

@@

\subsection{Parallel Algorithm on Distributed Memory Exploiting a Set of Evolvents}
\label{sub:parallel_evolvents}
\subsubsection{Rotated Evolvents}
One of the possible ways to overcome the negative effects of using a numerical
approximation of evolvent (it destroys the information about the neighbor points in
$\mathbb{R}^N$ space)
consists in using the multiple mappings
\begin{equation}
  Y_L(x)=\left\{y^0(x),\ y^1(x),...,\ y^L(x)\right\}
\end{equation}
instead of single Peano curve $y(x)$ (see \cite{strSergGO}).
The building of a set of Peano curves by rotation of the evolvents around the coordinate origin is a
distinctive feature was proposed in \cite{Gergel2009}. Taking into account the initial mapping, one
can conclude that current implementation of the
method allows to build up to $N(N-1)+1$ evolvents for mapping the $N$-dimensional domain
onto the corresponding one-dimensional intervals. This method for building a set of mappings can be
``scaled'' easily to obtain more evolvents (up to
$2^N$) if necessary.

Using the multiple mapping allows solving initial problem (\ref{eq:task}) by parallel solving the
problems
\[
\min\{\varphi(y^s(x)):x\in [0,1]\}, 1\leq s\leq S,
\]
on a set of intervals $[0,1]$ by the index method. Each one-dimensional problem is solved on a
separate processor. The trial results at the point \(x^k\) obtained for the problem being solved by
particular processor are interpreted as the results of the trials in the rest problems (in the
corresponding points \(x^{k_1},\dots,x^{k_S})\). In this approach, a trial at the point \(x^k \in
[0,1]\) executed in the framework of the \(s^{th}\) problem, consists in the following sequence
of
operations:
\par
1. Determine the image \(y^k=y^s (x^k)\) for the evolvent \(y^s (x)\).
\par
2. Inform the rest of processors about the start of the trial execution at the point \( y^k\) (the
blocking of the point \(y^k\) ).
\par
3. Determine the preimages \(x{}^{k_s}  \in [0,1], 1\leq s\leq S\), of the point \(y^k\) and
interpret the
trial executed at the point \(y^k \in D \) as the execution of the trials in the \(S\) points
\(x{}^{k_1} ,\dots,x{}^{k_s} \)
\par
4. Inform the rest of processors about the trial results at the point \(y^k\).
\par
The decision rules for the mentioned parallel algorithm, in general, are the same as the rules of the
sequential algorithm (except the method of the trial execution). Each processor has its own copy
of the software realizing the computations of the problem functions and the decision rule of the
index algorithm. For the organization of the interactions among the processors, the queues are
created on each processor, where the processors store the information on the executed iterations
in the form of the tuples: the processor number \(s\), the trial point \(x{}^{k_s}\).
The modifications of the method taking into account multiply criteria are given in \cite{GERGEL2017,Gergel2018}.
\par
The mentioned parallelization scheme was proposed in \cite{Gergel2009} and implemented in the
Globalizer system with the use of MPI technology. Main
features of implementation consist in the following:
\begin{itemize}
  \item A separate MPI-process is created for each
 of \(S\) one-dimensional problems being solved, usually, one process per one processor
 employed.
 \item Each process can use $p$ threads to parallel execute $p$ trials, usually one thread per an
accessible core.
\end{itemize}

\subsection{MIDACO Parallel Solver}
MIDACO (Mixed Integer Distributed Ant Colony Optimization) \cite{Schlueter2012} is a solver
which implements the evolutionary algorithm
the ant colony optimization metaheuristic for continuous search domains. ACO was originally
proposed to solve the TSP, but later has been successfully adopted to solve mixed integer nonlinear
programming problems \cite{SCHLUTER2009}.

The MIDACO solver utilizes reverse communication architecture \cite{Schlueter2012}.
Reverse communication means that the call of the objective function happens outside and
independently of the MIDACO source code.
Relying on this feature the following distributed parallelization scheme was implemented:
\begin{enumerate}
  \item MIDACO runs at the master node and generates \(n \times p\) new trial points at each
iteration.
  \item Each of \(n\) distributed computational nodes gets \(p\) points from the master and performs
\(p\) parallel trials on local computing devices.
  \item Each of \(n\) distributed computational nodes sends \(p\) values of the objective function to
the master.
\end{enumerate}

In this scheme all distributed communications were implemented using the MPI library.
Parallelism within a single node is powered by the OpenMP standard.

@@

\section{Results of Numerical Experiments}
\subsection{Sequential Algorithms}

@@

\subsection{Parallel Algorithms}

Computational experiments with parallel algorithms have been carried out on the Lobachevsky
supercomputer at State University of Nizhni Novgorod. A computational node includes 2 Intel
Sandy Bridge E5-2660 2.2 GHz processors, 64 GB RAM. Each CPU has 8 cores (i. e. total 16
cores were available per a node). All the computational nodes are working under CentOS Linux 7.2.
All the considered algorithms are implemented using C++ and compiled by the GCC 4.8.5.

The comparison of two parallel algorithms implemented in the Globalizer and MIDACO systems
was conducted on four-dimensional GKLS problems since the problems of larger dimensionality are
not supported by the free version of MIDACO. As in the case of the sequential algorithms, the
parameters of methods were determined for each problem class. For MIDACO, the parameter
\(focus = -1\). In the case of Globalizer, the parameter \(r\) was selected following Table
\ref{tab:params}. This parameter decreased by 0.3 with doubling the number of employed evolvents
since the increasing of the number of evolvents enhances the reliability of the method from Section
\ref{sub:parallel_evolvents} as well (see \cite{SOVRASOV2018} for details).

If one considers the costs of parallelism to be negligible as compared to the costs of computing the
objective functions in the optimization problems, the speedup in time due to the use of the parallel
method would be equal to the speedup with respect to the number of iterations. However, actually
this assumption is true if the computing of the objective function is well computation costly only. In
all numerical experiments, the time of computing the objective function was approximately $2.5\times
10^{-3}$ s.

In Table~\ref{tab:iterations}, an averaged number of iterations when solving 100 problems from
each considered class is presented.
The number of iterations is reduced considerably with increasing the number of nodes and the
number of threads on each node. Also, the number of solved problems in each class is given  in
brackets.
Globalizer solves almost all problems in both classes whereas MIDACO hasn't solved a considerable
number of problems from the GKLS 4d Hard class. Also, it is worth noting that the reliability of
MIDACO decreased with increasing number of nodes on both problem classes whereas the
reliability of Globalizer remains stable.

\begin{table}
  \centering
  \caption{Averaged numbers of iterations executed by the parallel algorithms for solving the test
optimization problems}
  \label{tab:iterations}
  \begin{tabular}{cccccccc}
    \cline{3-8}\noalign{\smallskip}
    \multicolumn{2}{c}{  } & \textit{p} & \multicolumn{2}{c}{Globalizer} & &
\multicolumn{2}{c}{MIDACO}   \\
    \noalign{\smallskip} \cline{4-5} \cline{7-8}  \noalign{\smallskip}
    \multicolumn{2}{c}{  } & & \textit{Simple} & \textit{Hard} & & \textit{Simple} &
\textit{Hard}  \\
    \noalign{\smallskip}\hline
    I & \textbf{1 cluster node}
      & \textit{1} &   25270 (100)  & 55180 (99) & & 27645 (98) & 72068 (71)  \\
    &  & \textit{16} & 1765 (100)  & 3714 (100) & &  1640 (97) & 4304 (70) \\
    \hline \noalign{\smallskip}
II  & \textbf{2 cluster nodes}  %\multirow{3}{*}{}
  & \textit{1} & 13056 (100) & 22938 (99) & & 10558 (89) & 27128 (73) \\
&   & \textit{16} & 732 (100) & 1759 (100)  &  & 1130 (92) & 2254 (73) \\
    \hline \noalign{\smallskip}
III & \textbf{4 cluster nodes} %\multirow{3}{*}{}
  & \textit{1}  & 5016 (100) & 12703 (100) & & 5777 (94) & 15980 (75) \\
& & \textit{16} & 367 (100) & 776 (100) & & 529 (88) &  1264 (66) \\
    \hline \noalign{\smallskip}
VI & \textbf{8 cluster nodes} %\multirow{3}{*}{}
  & \textit{1}  & 2103 (100) & 5063 (100) & & 2847 (97) & 9853 (83)\\
& & \textit{16} & 145 (100)  & 310 (100)  & & 272 (83) & 774 (57)\\
    \hline \noalign{\smallskip}
V & \textbf{12 cluster nodes} %\multirow{3}{*}{}
  & \textit{1}  & 1155 (100) & 2399 (100) & & 2233 (98) & 6022 (86) \\
& & \textit{16} & 76 (100)  & 159 (100)  & & 168 (87) & 393 (53)\\
    \noalign{\smallskip}\hline
  \end{tabular}
\end{table}

The averaged speedup in the number of iterations and the one in time (in brackets) are presented in
Table \ref{tab:speedup}. In the first row of Table \ref{tab:speedup}, the averaged number of
iterations and the averaged time of solving the problems in the sequential mode (in brackets) are
presented. MIDACO uses a smaller number of communications in the parallel computations than the Globalizer.
Therefore, the indicators of speedup increased monotonously with increasing number of
computational nodes employed. In Globalizer, each node should transfer the results of the executed
iterations to all other nodes. Therefore, the efficiency of parallelization decreased with increasing
number of nodes and the number of active threads on each node. This is evident from Table
\ref{tab:speedup} and Picture \ref{fig:speedup_p16}: the speedup in time decreased at the transition from 4 nodes to 8 ones when
using 16 threads. In the case of using 1 thread per a node, the speedups of both optimization methods
were approximately the same (see Picture \ref{fig:speedup_p16}). However, it should be noted that Globalizer executed less number of
trials, and spends much less time for solving all problems when \(p=1\). With increasing number of
nodes and threads, MIDACO approaches Globalizer in the averaged time of problem solving (see Pictures \ref{fig:time_p1}, \ref{fig:time_p16}),
however, its reliability decreases at the same time.

An anomalous superlinear speedup of MIDACO in time on the GKLS 2d Hard class when
employing 8 and 12 nodes can be explained by the fact that, according to Table \ref{tab:iterations},
MIDACO has solved a greater number of problems in the parallel mode than in the sequential one,
hence, has worked until the limit of trials is expired on a less number of problems.

\begin{figure}[ht]
  \centering
  \subfloat[$p=1$]{{\includegraphics[width=.5\textwidth]{images/speedup_midaco_globalizer_p_1.pdf}}\label{fig:speedup_p1}}
  \subfloat[$p=16$]{{\includegraphics[width=.5\textwidth]{images/speedup_midaco_globalizer_p_16.pdf}}\label{fig:speedup_p16}}
  \caption{Speedup demonstrated by the parallel algorithms when solving problems from the GKLS 4d Hard class}
\end{figure}

\begin{figure}[ht]
  \centering
  \subfloat[$p=1$]{{\includegraphics[width=.5\textwidth]{images/time_midaco_globalizer_p_1.pdf}}\label{fig:time_p1}}
  \subfloat[$p=16$]{{\includegraphics[width=.5\textwidth]{images/time_midaco_globalizer_p_16.pdf}}\label{fig:time_p16}}
  \caption{Averaged execution time of the parallel algorithms when solving problems from the GKLS 4d Hard class}
\end{figure}

\begin{table}
  \centering
  \caption{Speedup of parallel computations executed by the parallel algorithms}
  \label{tab:speedup}
  \begin{tabular}{cccccccc}
    \cline{3-8}\noalign{\smallskip}
    \multicolumn{2}{c}{  } & \textit{p} & \multicolumn{2}{c}{Globalizer} & &
\multicolumn{2}{c}{MIDACO}   \\
    \noalign{\smallskip} \cline{4-5} \cline{7-8}  \noalign{\smallskip}
    \multicolumn{2}{c}{  } & & \textit{Simple} & \textit{Hard} & & \textit{Simple} &
\textit{Hard}  \\
    \noalign{\smallskip}\hline
I  & \textbf{1 cluster node}  %\multirow{3}{*}{}
    & \textit{1}   & 25270 (27.3s) & 55180 (61.8s) & & 27645 (32.2s) & 72068 (132.5s)  \\
  &  & \textit{16} & 14.3(11.7) & 14.9(12.6)  & &  16.9(14.4) & 16.7(14.4) \\
  \hline \noalign{\smallskip}
II  & \textbf{2 cluster nodes}  %\multirow{3}{*}{}
  & \textit{1}      &   1.9(1.9) & 2.4(2.2)  & & 2.6(1.7) & 2.7(2.3) \\
  &   & \textit{16} & 34.5(20.4) & 31.4(18.8) & & 24.5(18.8) & 32.0(29.1) \\
    \noalign{\smallskip}\hline	\noalign{\smallskip}
III  & \textbf{4 cluster nodes}  %\multirow{3}{*}{}
& \textit{1}      & 5.0(4.6) & 4.3(4.1) & & 4.8(3.9) & 4.5(4.4) \\
&   & \textit{16} & 68.8(23.7) & 71.2(25.8) & & 52.3(32.9) & 57.0(50.2) \\
  \noalign{\smallskip}\hline	\noalign{\smallskip}
VI & \textbf{8 cluster nodes} %\multirow{3}{*}{}
& \textit{1}    & 12.0(8.7) & 10.9(8.3) & & 9.7(8.2)    & 7.3(9.0)    \\
& & \textit{16} & 174.1(4.7) & 177.8(5.3) & & 101.6(51.6) & 93.1(84.1) \\
    \noalign{\smallskip}\hline
    V & \textbf{12 cluster nodes} %\multirow{3}{*}{}
    & \textit{1}    & 21.9(11.4)  & 23.0(12.8)  & & 12.4(11.1)  & 12.0(14.8)  \\
    & & \textit{16} & 333.5(2.7) & 347.9(3.0) & & 164.5(82.0) & 183.2(129.7) \\
        \noalign{\smallskip}\hline
  \end{tabular}
\end{table}

@@

\section*{Acknowledgements}
The study was supported by the Russian Science Foundation, project No 16-11-10150.

% ---- Bibliography ----
%
\bibliographystyle{spmpsci}
\bibliography{bibliography}{}

>>>>

% This is samplepaper.tex, a sample chapter demonstrating the
% LLNCS macro package for Springer Computer Science proceedings;
% Version 2.20 of 2017/10/04
%
\documentclass[runningheads]{llncs}
%
\usepackage{graphicx}
\usepackage{amssymb}
\usepackage{subfig}
\usepackage{cite}

% Used for displaying a sample figure. If possible, figure files should
% be included in EPS format.
%
% If you use the hyperref package, please uncomment the following line
% to display URLs in blue roman font according to Springer's eBook style:
% \renewcommand\UrlFont{\color{blue}\rmfamily}

\begin{document}
%
\title{Comparison of several derivative-free global optimization algorithms\thanks{The study was supported by the Russian Science Foundation, project No 16-11-10150.}}
%
%\titlerunning{Abbreviated paper title}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Vladislav Sovrasov}
%
\authorrunning{V. Sovrasov}
%
\institute{Lobachevsky State University of Nizhni Novgorod, Russia \\
  \email{sovrasov.vlad@gmail.com}
}%
\maketitle              % typeset the header of the contribution
%

@@

@@

\section{Results of Numerical Experiments}

@@

@@

% ---- Bibliography ----
%
\bibliographystyle{splncs04}
\bibliography{bibliography}{}
%
